---
title: "TDI Project"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(zoo)
tinytex:::is_tinytex()

```

### How does weather explain New York shooting's rates?

**Objective**: For Data Incubator Project, I apply Bayesian Multilevel to model the rates of shooting in the City of Greater New York. While the second level of the model pools information across space (based on census data, the social and economic background), the first level simulate the trend and variability in response to the daily weather. This arrangement allows us to explore the variability of response with regard to the different social/economic background. If time permits, at the later stages of this investigation, I incorporate the effect of air quality (using AQI) to inform the rates of gun violence.

**Data**: For this analysis, I use multiple sources of information. This includes NYPD historical rate dataset (2006-2018), NCDC daily weather data, NYC census data (NYC open data and Department of City Planning). Later, I will include the AQI (Air Quality Index) to study the relevancy of air pollution to the occurrence of shooting. The time-span of observation considered in this analysis is 2006-2018. 


#### Preliminary Analysis: Temporal Varaiblity

```{r include=FALSE, message=FALSE, cache=FALSE}

setwd('D:/Saman/TDI')

lines = length(readLines("Data_NYPD.csv"))

input=read.csv("Data_NYPD.csv", header = TRUE, fill = TRUE, sep=',')
df = input[c(4:6,8:12,17:18)]; df$id = input

dates = input$OCCUR_DATE
dates = sapply(1:length(dates), 
               function(i) as.numeric(unlist(strsplit(gsub(" 0:00","",dates[i]),'/'))))

df$MONTH = dates[1,]; df$DAY = dates[2,]; df$YEAR = dates[3,]

times = as.character(input$OCCUR_TIME); 
times = sapply(1:length(times),function(i) as.numeric(unlist(strsplit(times[i],":"))))
df$HOUR = times[1,]
df$monyear <- as.yearmon(paste(df$YEAR, df$MONTH), "%Y %m")
df$dates = as.Date(paste(dates[3,],'-',dates[1,],'-',dates[2,],sep=''))

structured_df <- aggregate(STATISTICAL_MURDER_FLAG~BORO+monyear, as.data.frame(df),length)
sample_of_data=df[sample(100,100),c(1,2,4,5,6,9,10,17)]
colnames(sample_of_data) <- c("Borough","Precinct","Murder","Age Group","Sex","Lat","Lon","Dates")

```

A sample of NYDP shooting dataset looks like the following:

```{r  message=FALSE, cache=FALSE}
head(sample_of_data,4)

```

During recent years, the New York shooting rates constantly decreased. This negative trends in year-to-year data can root from diffrent sources (such as increase in the number of police stations, education and etc.) 


```{r include=FALSE, message=FALSE, cache=FALSE}

library(reshape2)
library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)

df_plot <- structured_df[which(structured_df[,1]=='BROOKLYN'),]
df_plot <- melt(df_plot[, c("monyear", "STATISTICAL_MURDER_FLAG")], id="monyear")
p1=ggplot(df_plot) + geom_line(aes(x=monyear, y=value),color='black',size = 1.2)+ 
  stat_smooth(aes(x = monyear, y = value),color='red', method = "lm",
              formula = y ~ poly(x, 1), se = TRUE)+ ggtitle("BROOKLYN")+
  labs(x = "Date", y='Shooting')

df_plot <- structured_df[which(structured_df[,1]=='BRONX'),]
df_plot <- melt(df_plot[, c("monyear", "STATISTICAL_MURDER_FLAG")], id="monyear")

p2=ggplot(df_plot) + geom_line(aes(x=monyear, y=value),color='black',size = 1.2)+ 
  stat_smooth(aes(x = monyear, y = value),color='red', method = "lm",
              formula = y ~ poly(x, 1), se = TRUE)+ ggtitle("BRONX")+
  labs(x = "Date", y='Shooting')

df_plot <- structured_df[which(structured_df[,1]=='MANHATTAN'),]
df_plot <- melt(df_plot[, c("monyear", "STATISTICAL_MURDER_FLAG")], id="monyear")

p3=ggplot(df_plot) + geom_line(aes(x=monyear, y=value),color='black',size = 1.2)+ 
  stat_smooth(aes(x = monyear, y = value),color='red', method = "lm",
              formula = y ~ poly(x, 1), se = TRUE)+ ggtitle("MANHATTAN")+
  labs(x = "Date", y='Shooting')

df_plot <- structured_df[which(structured_df[,1]=='QUEENS'),]
df_plot <- melt(df_plot[, c("monyear", "STATISTICAL_MURDER_FLAG")], id="monyear")

p4=ggplot(df_plot) + geom_line(aes(x=monyear, y=value),color='black',size = 1.2)+ 
  stat_smooth(aes(x = monyear, y = value),color='red', method = "lm",
              formula = y ~ poly(x, 1), se = TRUE)+ ggtitle("QUEENS")+
  labs(x = "Date", y='Shooting')


```
```{r, fig.width=8, fig.height=4,message=FALSE,}

grid.arrange(p1, p2, p3, p4, ncol = 2)

```


However, the shooting rates present a seasonal pattern, suggesting a correlation with weather variability. The Following figure shows that the number of incidents in the warmer months is higher than the shooting rates in colder seasons.

```{r, fig.width=8, fig.height=4}

df_plot <- aggregate(STATISTICAL_MURDER_FLAG~MONTH,as.data.frame(df),length)
barplot(sapply(1:12,function(x) length(df$STATISTICAL_MURDER_FLAG[which(df$MONTH==x)])),
        col=rainbow(12),names=month.abb[1:12],ylab='Number of Shootings',
        font=2,cex.axis = 0.7,cex.names=0.7,cex.lab=0.8)

```



```{r include=FALSE, message=FALSE, cache=FALSE}

weather_data=read.csv('D:/Saman/TDI/weather.csv')
weather_dates=as.numeric(as.Date(weather_data$DATE))

structured_df <- aggregate(STATISTICAL_MURDER_FLAG~dates, as.data.frame(df),length)
id_dates_shooting=as.numeric(as.Date(structured_df$dates))

id_dates=as.numeric(as.Date(df$dates))

ranks=apply(weather_data[,c(4,7:9)],2,function(x) round(rank(x)/length(x),2))

weather_shooting=sapply(1:length(id_dates_shooting), 
       function(x) ranks[which(weather_dates==id_dates_shooting[x]),])


thresh=quantile(structured_df$STATISTICAL_MURDER_FLAG,seq(0,1,0.2))
thresh=as.numeric(thresh)+0.5
labels_ins <- c('0-0.2','0.2-0.4','0.4-0.6','0.6-0.8','0.8-1')

classified_shootings=cut(structured_df$STATISTICAL_MURDER_FLAG,
       breaks = as.numeric(thresh), 
       labels = as.character(1:(length(thresh)-1)))

```


The additional evidence is found in comparing the daily shooting incidents with weather variables. On the daily number of incidents, I extract 6 different classes with different range of quantiles. Further, I compare the outcome with the quantile of the corresponding maximum daily temperature. The boxplots suggest that the higher number of shooting correlates with higher temperature.

```{r, fig.width=8, fig.height=4.5,echo=FALSE}

boxplot(sapply(1:(length(thresh)-1), function(x) weather_shooting[3,which(classified_shootings==x)]),col='grey',
        staplecol=rainbow(5),whiskcol=rainbow(5),
        boxcol=rainbow(5),medcol='black', 
        xlab='Quantile Ranges of Shoorting Rates',
        whisklty=1,ylab='Max Temperature Quantile',
        outline=FALSE,names=labels_ins,boxwex = 0.3,
        names.font=2,cex.axis = 0.7,cex.names=0.7,cex.lab=0.8)

```

```{r, include=FALSE, fig.width=8, fig.height=3.2,echo=FALSE}

boxplot(sapply(1:(length(thresh)-1), function(x) weather_shooting[1,which(classified_shootings==x)]),
        staplecol=rainbow(5),whiskcol=rainbow(5),
        boxcol=rainbow(5),medcol=rainbow(5),
        xlab='Quantile Ranges of Shoorting Rates',
        whisklty=1,ylab='Precipitation Quantile',
        outline=FALSE,names=labels_ins,boxwex = 0.4,
        font=2,cex.axis = 0.7,cex.names=0.7,cex.lab=0.8)

```

#### Preliminary Analysis: Spatial Varaiblity

The spatial distribution of shooting incident yields useable information on how the affected areas changed over time. The following map demonstrates the location of gun violence in 2007.

#####

```{r fig.width=8, fig.height=5,message=FALSE, cache=FALSE, echo=FALSE}

library(ggmap)

structured_df <- aggregate(STATISTICAL_MURDER_FLAG~BORO+YEAR+Latitude+Longitude, as.data.frame(df),length)
plotvar = structured_df[which(structured_df[,2]==2007),]
qmplot(Longitude, Latitude, data = plotvar, maptype = "toner-lite", color = I("red"),margins=margin)
```

#####

Using the Kernel estimators, I looked at the density of the number of incindents within 2-km radius. It is observed that the density of crimes decrease over time. 

```{r include=FALSE, message=FALSE, cache=FALSE}

library(pointdensityP)
library(ggmap)

register_google(key = "AIzaSyAtN-yFKd2hMFh1Td-lIumwj_57I5ZcQXg")
has_google_key()

density_shot = NULL

for (i in 1:12){
  
  yr_ind = 2006:2018
  
  dp=pointdensity(df[which(df$YEAR==yr_ind[i]),], lat_col="Latitude", lon_col="Longitude",date_col = "dates", 1, 2)
  density_shot = c(density_shot,max(dp$count))
  nyc_map <- get_map(location = c(lon = -73.95, lat = 40.73), maptype = "terrain", zoom = 11)
  p=ggmap(nyc_map)
  p=p + geom_point(data=dp[c((nrow(dp)-40),nrow(dp)),], aes(x=lon, y=lat),size=12,color='darkred', alpha=0.5)+ theme(axis.text=element_text(size=6),
        axis.title=element_text(size=10,face="bold"))+ggtitle(yr_ind[i])
  assign(paste("P", i, sep = ""), p)  
}

pcomb=grid.arrange(P3, P6, P9, P12, ncol = 2)

```

```{r, fig.width=8, fig.height=2.5,echo=FALSE}

scaleFUN <- function(x) sprintf("%.0f", x)

df_plot = list(); df_plot$Density = density_shot; df_plot$Year=c(2005.5:2016.5)
df_plot = as.data.frame(df_plot)
ggplot(df_plot) + geom_point(aes(x=Year, y=Density),color='black',size = 3)+stat_smooth(aes(x = Year, y = Density), method = "lm",
              formula = y ~ poly(x, 3), se = FALSE) +
  ggtitle("Density")+scale_x_continuous(labels=scaleFUN)

```

Moreover, the following maps indicate the top-40 highest density points throughout the years. We can see that the highest density point of crimes from Brooklyn in 2009 shifts to the Bronx in 2014. In this project, I plan to study this shift and relate that to the social/economic background of the neighborhood.  I ask as to what extent the change in latter, modulate the spatial shift in the crime rates.

```{r, fig.width=8, fig.height=8,message=FALSE, echo=FALSE}

grid.newpage()
grid.draw(pcomb)

```

